---
title: "5501-01 programming assignment"
author: "Steve Simon and Leroy Wheeler"
format: 
  html:
    embed-resources: true
date: 2024-08-18
---

This program reads data on housing prices in Albuquerque, New Mexico in 1993. Find more information in the [data dictionary][dd].

[dd]: https://github.com/pmean/datasets/blob/master/albuquerque-housing.yaml

This code is placed in the public domain.

## Load the tidyverse library

For most of your programs, you should load the tidyverse library. The messages and warnings are suppressed.

```{r setup}
#| message: false
#| warning: false
library(tidyverse)
```

## Read the data and view a brief summary

Use the read_csv function to read the data. The glimpse function will produce a brief summary.

```{r read}
alb <- read_csv(
  file="../data/albuquerque-housing.csv",
  col_names=TRUE,
  col_types="nnnnccc",
  na=".")
glimpse(alb)
```

## Calculate overall means

The summarize_if function produces means, but only for numeric data. You wouldn't want to compute means for data with values "yes" and "no".

```{r means}
alb |>
  summarise_if(is.numeric, mean, na.rm = TRUE)
```

## Summarize price

The average price of a home, 106 thousand dollars, is quite low because the data comes from 1993.

## Summarize sqft

The average size of a house is about 1,600 square feet, which is small by today's standards.

## Summarize age

The average age of a home was 15 years, which is a relatively new house.

## Summarize features

The average number of features is 3.5, which is low.