---
title: "Sample size justification for postural sway study"
format: 
  html:
    embed-resources: true
editor: source
---

This program evaluates various sample size calculations for a proposed replication of the postural-sway study in a different population. It was written by Steve Simon and Leroy Wheeler on 2024-10-09 and is placed in the public domain.

## Scenario 1

-   Replicate postural sway study
    -   Different populations
    -   Same outcome measure
-   Research hypothesis, $H_0\ \mu_1 - \mu_2 = 0$
-   Standard deviations: 9.77, 4.09
-   MCID = 4

```{r power-1}
power.t.test(
  n=NULL,
  delta=4,
  sd=9.8,
  sig.level=0.05,
  power=0.9,
  type="two.sample",
  alternative="two.sided")
```

With a sample of 128 patients, we would have 90% power for detecting a 4 unit difference in postural sway, using a two-sided test at an alpha level of 0.05.

## Scenario 2, MCID = 2

```{r power-2}
power.t.test(
  n=NULL,
  delta=2,
  sd=9.8,
  sig.level=0.05,
  power=0.9,
  type="two.sample",
  alternative="two.sided")
```

If we wanted to be able to detect a 2 unit difference, we would need a sample size that is about four times as large, assuming standard deviation, our alpha and power were unchanged.
    
## Scenario 3, MCID=8

```{r power-3}
power.t.test(
  n=NULL,
  delta=8,
  sd=9.8,
  sig.level=0.05,
  power=0.9,
  type="two.sample",
  alternative="two.sided")
```
With a sample of 33 patients, we would have 90% power for detecting an 8 unit difference in postural sway, using a two-sided test at an alpha level of 0.05.

## Scenario 4, sd=4.9

```{r power-4}
power.t.test(
  n=NULL,
  delta=4,
  sd=4.9,
  sig.level=0.05,
  power=0.9,
  type="two.sample",
  alternative="two.sided")
```
With a sample of 33 patients, we would have 90% power for detecting a 4 unit difference in postural sway with a standard deviation of 4.9, using a two-sided test at an alpha level of 0.05.


## Scenario 5, sd=19.6

```{r power-5}
power.t.test(
  n=NULL,
  delta=4,
  sd=19.6,
  sig.level=0.05,
  power=0.9,
  type="two.sample",
  alternative="two.sided")
```
If we increased our pooled standard deviation from 9.8 to 19.6, we would need 506 patients to have 90% power for detecting a 4 unit difference in postural sway, using a two-sided test at an alpha level of 0.05.

## Scenario 6, alpha=0.01

```{r power-6}
power.t.test(
  n=NULL,
  delta=4,
  sd=9.8,
  sig.level=0.01,
  power=0.9,
  type="two.sample",
  alternative="two.sided")
```
With a sample of 181 patients, we would have 90% power for detecting a 4 unit difference in postural sway with a standard deviation of 9.8, using a two-sided test at an alpha level of 0.01.

## Scenario 7, alpha=0.10

```{r power-7}
power.t.test(
  n=NULL,
  delta=4,
  sd=9.8,
  sig.level=0.1,
  power=0.9,
  type="two.sample",
  alternative="two.sided")
```
With a sample of 104 patients, we would have 90% power for detecting a 4 unit difference in postural sway with a standard deviation of 9.8, using a two-sided test at an alpha level of 0.1.

## Scenario 8, power=0.8

```{r power-8}
power.t.test(
  n=NULL,
  delta=4,
  sd=9.8,
  sig.level=0.05,
  power=0.8,
  type="two.sample",
  alternative="two.sided")
```
With a sample of 96 patients, we would have 80% power for detecting a 4 unit difference in postural sway with a standard deviation of 9.8, using a two-sided test at an alpha level of 0.05.

## Scenario 9, power=0.95

```{r power-9}
power.t.test(
  n=NULL,
  delta=4,
  sd=9.8,
  sig.level=0.05,
  power=0.95,
  type="two.sample",
  alternative="two.sided")
```
With a sample of 157 patients, we would have 95% power for detecting a 4 unit difference in postural sway with a standard deviation of 9.8, using a two-sided test at an alpha level of 0.05.