---
title: "Normal probabilities and quantiles"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
execute: 
  error: true
---

This program displays the standard normal curve along with various probabilities and quantiles. It was written by Steve Simon and Leroy Wheeler on 2024-09-04 and is placed in the public domain.

## Load the tidyverse library

```{r setup}
#| message: false
#| warning: false
library(tidyverse)
```

## Using R to draw the standard normal curve

use seq to calculate 100 evenly spaced values between -4 and +4 and dnorm to compute the bell curve at each point. Use geom_polygon to paint the area surrounded by the bell curve.

```{r standard-normal}
x <- seq(-4, 4, length=100)
y <- dnorm(x)
data.frame(x, y) |>
  ggplot(aes(x, y)) + 
    geom_polygon(fill="white", color="black") -> normal_curve
normal_curve
```

## P[-1 < Z < 1]

```{r homework-1}
a <- 1
normal_curve +
  geom_vline(xintercept=-a) +
  geom_vline(xintercept= a) +
  geom_label(x=-a, y=0.4, label=-a) +
  geom_label(x= a, y=0.4, label= a) +
  geom_label(x=0, y=0, label="Area = ?")
pnorm(1) - pnorm(-1)
```
There is a 68% percent chance of getting values between 1 and -1 in the above standard normal distribution curve.


## 60th percentile of a standard normal

Use qnorm to calculate quantiles of the standard normal distribution.

```{r quantile homework-2}
p <- 0.6
a <- qnorm(p)
normal_curve +
  geom_vline(xintercept=a) +
  geom_label(x=a, y=0.4, label="Quantile = ?") +
  geom_label(x=a-0.5, y=0, label=p)
qnorm(0.6)
```
The value of 0.25 on a standard normal distribution curve is the 60th percentile.  This means that there is a 60% probability of getting values less than 0.25 and 40% probability of getting values greater than 0.25.