---
title: "Analysis of Titanic dataset"
format: 
  html:
    embed-resources: true
---

This program reads data on survival of passengers on the Titanic. Find more
information in the [data dictionary][web00].

[web00]: https://github.com/pmean/data/blob/main/files/titanic.yaml

This code was written by Steve Simon and Leroy Wheeler on 2024-11-13 and is placed in the public
domain.

## Load the tidyverse library

```{r}
#| label: setup
#| message: false
#| warning: false
library(broom)
library(epitools)
library(tidyverse)
```


## Read the data and view a brief summary

```{r}
#| label: read
ti <- read_tsv(
  file="../data/titanic.txt",
  col_names=TRUE,
  col_types="ccncn",
  na="NA")
names(ti) <- tolower(names(ti))
glimpse(ti)
```

## Replace numeric codes for survived

```{r}
#| label: replace-numbers
ti$survived <-
	factor(
		ti$survived,
		level=1:0,
		labels=c("yes", "no"))
```


## Question 1: Create a new variable, third_class that indicates whether a passenger is in third class or not.
```{r}
#| label: third_class
ti$third_class <- 
  case_when(
    ti$pclass == "1st" ~ "no",
    ti$pclass == "2nd" ~ "no",
    ti$pclass == "3rd" ~ "yes")
```

## Question 2: What are the probabilities of survival for third class passengers. How does this compare to the probability of survival for the other passengers.

## Get counts of third class passengers by survival

```{r}
#| label: counts
table1 <-xtabs(~third_class+survived, data=ti)
table1
```
## Get proportions for died/survived by third class status

```{r}
#| label: proportions
table1 |>
  proportions("third_class")
```
## Interpretation of the output for question 2.

If you were a third class passenger, you had less than a 20% chance of survival, whereas if you were a first or second class passenger, you had more than a 50% chance of survival.

## Question 3: Test the hypothesis that the survival probability is different for third class passengers and the other passengers. Interpret the p-value and confidence interval.

```{r}
#| label: equality-of-proportions
prop.test(table1, correct=FALSE)
```
## Interpretation of the output for question 3. 

The Chi-squared statistic is much larger than the degrees of freedom and the
p-value is small. Therefore we will reject the null hypothesis and conclude that 
there is a statistically significant difference in the mortality rates between
the third class passengers and the 1st/2nd class passengers.
The 95% confidence interval for the difference in proportions is 0.27 to 0.37.
This interval excludes the value of zero and indicates that the mortality rate is
at least 27% higher and possibly as much as 37% higher for third class passengers
compared to the rest of the 1st/2nd class passengers.