---
title: "Analysis of gardasil shots by demographic factors"
format: 
  html:
    embed-resources: true
---

This program reads data on Gardasil vaccinations in young women. Find more information in the [data dictionary][dd].

[dd]: https://raw.githubusercontent.com/pmean/datasets/master/gardasil.yaml

The program was written by Steve Simon on 2024-09-07 and is placed in the public domain.

## Load the tidyverse library

For most of your programs, you should load the tidyverse library. The messages and warnings are suppressed.

```{r setup}
#| message: false
#| warning: false
library(tidyverse)
```

## Read the data and view a brief summary

Use the read_csv function to read the data. The glimpse function will produce a brief summary. Use tolower to convert uppercase to lowercase.

```{r read}
gard <- read_csv(
  file="../data/gardasil.csv",
  col_names=TRUE,
  col_types="nnnnnnnnnn")
names(gard) <- tolower(names(gard))
glimpse(gard)
```

## Question 7: First create factors for medassist

The factor function identifies a variable as categorical and assigns labels to number codes. You don't necessarily need to use factor if the data you read in is character strings, as R automatically treats those variable as categorical.

```{r medassist-1}
gard$medassist <- factor(
  gard$medassist,
  levels=0:1,
  labels=c(
    "No medical assistance",
    "Received medical assistance"))
```

## Question 7: Summarize and interpret the percentage of patients receiving medical assistance. Be sure to convert the number codes for this variable into labels using the factor function

```{r medassist-2}
gard |> 
  count(medassist) |>
  mutate(total=sum(n)) |>
  mutate(pct=round(100*n/total))
```

Eighty one percent of patients received at least some medical assistance while the remaining 19% did not.

## Create factors for shots

It is a bit silly to replace 1, 2, 3 with One, Two, Three. The main reason is to clearly identify shots as categorical rather than continuous.

```{r shots-1}
gard$shots <- factor(
  gard$shots,
  levels=1:3,
  labels=c(
    "One", 
    "Two", 
    "Three"))
```

## Counts and percentages for shots

```{r shots-2}
gard |> 
  count(shots) |>
  mutate(total=sum(n)) |>
  mutate(pct=round(100*n/total))
```

Slightly more patients got three shots than one or two shots, but this is still less than half of the patients overall.

## Question 8: First calculate the percentages for number of shots received by whether the patient received medical assistance. Interpret this chart. 

```{r shots-by-medassist-1}
gard |>
  count(medassist, shots) |>
  group_by(medassist) |>
  mutate(row_total=sum(n)) |>
  mutate(pct=round(100*n/row_total))
```
Surprisingly 41% of patients who did not receive medical assistance received all three shots when compared to the 25% of patients who received medical assistance.


## ## Question 8: Draw a bar chart showing the percentages for number of shots received by whether the patient received medical assistance. Interpret this chart.

```{r shots-by-medassist-2}
gard |>
  ggplot(aes(x=medassist, fill=shots)) +
    geom_bar(position="fill") +
    xlab("Med assist") +
    ylab("Proportion") +
    ggtitle("Plot drawn by Leroy Wheeler on 2024-09-12")
```

Patients who did not receive medical assistance were more likely to complete the full round of three Gardisil shots compared to patients who received some medical assistance.

