---
title: "Analysis of Titanic dataset Module 14"
format: 
  html:
    embed-resources: true
---

This program reads data on survival of passengers on the Titanic. Find more
information in the [data dictionary][dd].

[dd]: https://github.com/pmean/data/blob/main/files/titanic.yaml

This code was written by Steve Simon and Leroy Wheeler on 2024-11-19 and is placed in the public
domain.

## Load the tidyverse library

```{r}
#| label: setup
#| message: false
#| warning: false
library(epitools)
library(tidyverse)
```

## Read the data and view a brief summary

```{r}
#| label: read
ti <- read_tsv(
  file="../data/titanic.txt",
  col_names=TRUE,
  col_types="ccncn",
  na="NA")
names(ti) <- tolower(names(ti))
glimpse(ti)
```

## Replace numeric codes for survived

```{r}
#| label: replace-numbers
ti$survived <-
	factor(
		ti$survived,
		level=1:0,
		labels=c("yes", "no"))
```

## Question 1: Create a new variable, third_class that indicates whether a passenger is in third class or not. What is the odds ratio comparing survival between third class passengers and first/second class passengers? Interpret this odds ratio and the associated confidence interval.

```{r}
#| label: third_class
ti$third_class <- 
  case_when(
    ti$pclass == "1st" ~ "no",
    ti$pclass == "2nd" ~ "no",
    ti$pclass == "3rd" ~ "yes")
```


## Get counts of sex by passenger class

```{r}
#| label: counts
table1 <-xtabs(~third_class+survived, data=ti)
table1
```
## Odds ratio calculation

```{r}
#| label: odds-ratio
oddsratio(table1)
```

#### Interpretation of the odd ratio output

We are 95% confident that the odds ratio of survival for 1st/2nd class passengers
is at least 3.5 and possibly as large as 5.7, after accounting for sampling error. 
This interval excludes the value of 1, so we conclude that the odds of 
survival is 4.5 times higher for 1st and 2nd class passengers than for third class passengers.

## Question 2: Calculate a chi-squared test of independence that examines the association between passenger class (third versus first/second) and mortality. Interpret the test result.

```{r}
#| label: chi-squared-test-1
m1 <- chisq.test(table1, correct=FALSE)
m1
```

#### Interpretation of the chi-square output

Because the chi-squared statistic is much larger than the degrees of freedom and the
p-value is small, we will reject the null hypothesis and conclude that passenger class status
and survival are related (not independent)
