---
title: "Analysis of relationships in pulmonary data"
format: 
  html:
    embed-resources: true
editor: source
---

This program assesses the relationships among variables in a study of pulmonary function in children. There is a [data dictionary][dd] that provides more details about the data. The program was written by Steve Simon on 2024-09-07 and is placed in the public domain.

[dd]: https://github.com/pmean/datasets/blob/master/fev.yaml

## Libraries

The tidyverse library is the only one you need for  this program.

```{r setup}
#| message: false
#| warning: false
library(tidyverse)
```

## List variable names

Since the variable names are not listed in the data file itself, you need to list them here.

```{r names}
pulmonary_names <- c(
    "age",
    "fev",
    "ht",
    "sex",
    "smoke")
```

## Reading the data

Here is the code to read the data and show a glimpse. 

```{r read}
pulmonary <- read_csv(
  file="../data/fev.csv",
  col_names=pulmonary_names,
  col_types="nnncc")
glimpse(pulmonary)
```

## Question 1: Update the program to calculate descriptive statistics (mean, standard deviation, minimum, and maximum) for ht. Interpret these statistics.

```{r descriptive-ht}
summary(pulmonary$ht)
sd(pulmonary$ht)
```

The mean height is about 61 inches with a standard deviation of almost 6 inches. Height ranges from 46 to 74 inches, which is consistent with a pediatric population.


## Question 2: Draw a scatterplot of ht versus fev. Place ht on the x-axis and fev on the y-axis. Interpret this plot.


```{r plot-ht-fev}
pulmonary |>
  ggplot(aes(ht, fev)) +
    geom_point() +
    geom_smooth() +
    xlab("Height (inches)") +
    ylab("Forced Expiratory Volume (Liters)") +
    ggtitle("Scatter plot drawn by Leroy Wheeler on 2024-09-11")
```

There is a positive linear association between height and fev. Calculation of r will likely confirm this observation.



## Question 3: Calculate the correlation between ht and fev. Interpret this correlation.
```{r correlation}
cor(pulmonary$ht, pulmonary$fev)
```
A correlation value of r=0.87 confirms the strong positive relationship between height and fev in this data set.

## Question 4: Calculate counts and percentages for sex. Please be sure to convert sex from the numeric codes into a factor. Interpret these statistics.

```{r descriptive-sex}
pulmonary |>
  count(sex) |>
  mutate(total=sum(n)) |>
  mutate(pct=round(100*n/total))
```
The data set was roughly split in half according to sex with 51% males and 49% females.

## Question 5: Draw a boxplot for sex and fev. Interpret this boxplot
```{r plot-sex-fev}
pulmonary |>
  ggplot(aes(sex, fev)) +
    geom_boxplot() +
    xlab("Sex (F/M)") +
    ylab("Forced Expiratory Volume (Liters)") +
    ggtitle("Box plot drawn by Leroy Wheeler on 2024-09-11")
```

The fev values are a little larger for males when compared to females.  The variability for the male data is also slightly higher as well. These results are not surprising.

## Question 6: Calculate the difference in average fev values between males and females. Is this a large or a small difference? Calculate the effect size by dividing by the standard deviation of the females. Is this a small, medium, or large effect size?

```{r mean-fev-by-sex}
pulmonary |>
  group_by(sex) |>
  summarize(
    mean_fev=mean(fev),
    sd_fev=sd(fev))
```

The average fev values for males is 2.8 which is larger than that observed in females, which is 2.5. Males also have a standard deviation of 1, which is also higher than the standard deviation of 0.6 seen in females.

The effect size between males and females is approximately 0.6 standard deviations.
